Choosing the Choice: Reflections on Modelling Decisions and Behaviour in Demographic Agent-Based Models

### 2017-03-26 article.tex
review comments A2,3,4 - first pass

We suggest that inclusion and operationalisation of context are best tackled through an iterative process of development, and evaluation against empirical data, somewhat similar to that proposed by \cite{Cioffi-Revilla2010}. This approach places considerable demands on the modeller in terms of developing the model, but also in exercising their judgement as to what should be included, and identifying where a component should be iterated out. This is however an expensive exercise, both because augmenting a model is time consuming, and because evaluating complex models poses considerable challenges. The challenge of evaluation can be somewhat ameliorated by the use of techniques like sensitivity analysis (see \citet{Thiele2014} for a review of several techniques, and \citet{Oakley2004} for an alternative approach), which can support the modeller in identifying how the individual parameters of a model contribute to overall outputs. A key benefit to this is that parameters which do not significantly impact the results of a model can be removed. This too has limitations, because as we discuss later in relation to heterogeneity in agents (section \ref{heterogeneity-in-decision-strategies}), context can take the form of both process, and parameter, and sensitivity analysis is directly informative only about the latter.
As a result, the onus remains on the modeller to use their best judgement.

@@

Time is explicitly represented in almost all agent-based models, either in discrete or (pseudo) continuous form, in order to allow the dynamic unfolding of the modelled processes by `stepping' from one time-step to the next (or between time-ordered events in continuous-time discrete-event simulation). However, time often  does not enter into the decision making process of the simulated individuals, in that agents do not differ in their preferences between rewards realised immediately and equivalent rewards realised at some point in the future.

@@

All of these problems are far from trivial, particularly as there is little consensus about which models of decision are best in which contexts. How then should we proceed with modelling when the fundamental questions about how individuals make decisions, how they treat time, and how they differ between themselves in these factors are still very uncertain? One way forward is to use the same simulation set-up to analyse multiple models of behaviour \citep{Epstein2013, Rossiter2014}. This multi-model approach allows the modeller to attempt to distinguish between more and less plausible models of behaviour in the particular research context. Conditional on the simulation set-up being appropriate, more plausible models may be better able to reproduce multiple empirical patterns at varying scales \citep{Werker2004, Bianchi2008}.

This commitment to analysing multiple models of behaviour within a single simulation project is facilitated by a modular approach to simulation design, so that different models of behaviour can be swapped in and out without changing the underlying conditions of the simulation \citep{Epstein2013}, as can be seen in \citet{Gray2016}. Such an approach reflects standard object-oriented design principles, but is not always reflected in an academic context (see also \citet{Rossiter2015} for discussion of the application of software engineering principles to social simulation). Modularity also allows for the possibility of easy extension by other researchers with access to the code, allowing for further behavioural models to be examined within the same simulation framework.

So, in the light of the above arguments, how to choose the appropriate representation of the human decisions and behavior to be used in the agent-based modelling of demographic processes? First, we need to remember that agent-based models are a very convenient tool, with which we can integrate behavioural theory with social theory and data. This implies that – as suggested in the Manifesto of computational social science \citep{Conte} – such models would ideally need to collect additional information on human decision making through bespoke cognitive experiments, instead of merely relying on hypotheses or assumptions \citep{Courgeau}. In this way, the choice of a particular choice model will become an empirical issue, but specific to a given context and the decision problem of interest. Following this approach would allow for including empirical micro-foundations in computational demographic models in an open and transparent way – and provide a natural, bottom-up and empirical way if evaluating the suitability of particular choice theories for the problem at hand.

>>>>

The problems of inclusion and operationalisation of context are best tackled through an iterative process of development and evaluation against empirical data, in agreement with proposals from \cite{Cioffi-Revilla2010} and from the ecological literature on Agent-Based Models \citep{Grimm2005, Schmolke2010}. The Pattern-Oriented Modelling approach of \cite{Grimm2005} describes a process of matching empirical patterns at multiple levels in order to determine the suitability of the model under study. \cite{Cioffi-Revilla2010} recommends a process of movement from simple to complex models, where at each stage an additional component of the phenomenon under study can be added, until a satisfactory model is found.

This iterative approach places considerable demands on the modeller in terms of developing the model, but also in exercising their judgement as to what should be included, and identifying where a component should be iterated out. This is however an expensive exercise, both because augmenting a model is time consuming, and because evaluating complex models poses considerable challenges. The challenge of evaluation can be somewhat ameliorated by the use of techniques like sensitivity analysis (see \citet{Thiele2014} for a review of several techniques, and \citet{Oakley2004} for an alternative approach), which can support the modeller in identifying how the individual parameters of a model contribute to overall outputs. A key benefit to this is that parameters which do not significantly impact the results of a model can be removed. This too has limitations, because as we discuss later in relation to heterogeneity in agents (section \ref{heterogeneity-in-decision-strategies}), context can take the form of both process, and parameter, and sensitivity analysis is directly informative only about the latter.
As a result, the onus remains on the modeller to use their best judgement. The application of such judgement can be documented, however: \cite{Schmolke2010} suggest that developmental cycles are recorded in a standardised format, the TRACE framework, so the process of rejection of alternatives through which the final model came about can be reconstructed.

@@

Time is explicitly represented in almost all agent-based models, either in discrete or (pseudo) continuous form, in order to allow the dynamic unfolding of the modelled processes by `stepping' from one time-step to the next (or between time-ordered events in continuous-time discrete-event simulation). However, time often  does not enter into the decision making process of the simulated individuals. This results in the exclusion of a number of factors that may potentially be significant in eventual model outcomes. Firstly, individuals do not always make decisions instantaneously; as \cite{Willekens2017} describe elsewhere in this special issue, the theory of planned behaviour can be used to describe the gradual process of intention forming, influence-absorption, information gathering and planning that terminate in a migration event. This drawn-out decision making process allows the possibility that agents may change their minds. Secondly, past experiences affect decisions in the present. This brings up questions about how the passage of time affects memory and attitudes. Finally, agents differ in their preferences between rewards realised immediately and equivalent rewards realised at some point in the future.

@@

All of these problems are far from trivial, particularly as there is little consensus about which models of decision are best in which contexts. How then should we proceed with modelling when the fundamental questions about how individuals make decisions, how they treat time, and how they differ between themselves in these factors are still very uncertain? One way forward is to use the same simulation set-up to analyse multiple models of behaviour \citep{Grimm2005, Epstein2013, Rossiter2014}. This multi-model approach allows the modeller to attempt to distinguish between more and less plausible models of behaviour in the particular research context. Conditional on the simulation set-up being appropriate, more plausible models may be better able to reproduce multiple empirical patterns at varying scales \citep{Werker2004, Bianchi2008}.

All of these problems are far from trivial, particularly as there is little consensus about which models of decision are best in which contexts. How then should we proceed with modelling when the fundamental questions about how individuals make decisions, how they treat time, and how they differ between themselves in these factors are still very uncertain? One way forward is to use the same simulation set-up to analyse multiple models of behaviour \citep{Grimm2005, Epstein2013, Rossiter2014}. This multi-model approach allows the modeller to attempt to distinguish between more and less plausible models of behaviour in the particular research context . Conditional on the simulation set-up being appropriate, more plausible models may be better able to reproduce multiple empirical patterns at varying scales \citep{Werker2004, Bianchi2008}.

This commitment to analysing multiple models of behaviour within a single simulation project is facilitated by a modular approach to simulation design, so that different models of behaviour can be swapped in and out without changing the underlying conditions of the simulation \citep{Epstein2013}, as can be seen in \citet{Gray2016}. Such an approach reflects standard object-oriented design principles, but is not always reflected in an academic context (see also \citet{Rossiter2015} for discussion of the application of software engineering principles to social simulation). Modularity also allows for the possibility of easy extension by other researchers with access to the code, allowing for further behavioural models to be examined within the same simulation framework. On this theme, \cite{Bell2015} highlight the atomistic nature of the agent-based modelling discipline, and advocate the introduction of Agent-based Modelling Primitives (AMPs) to enable general components of agent-based models to be packaged and re-used.

So, in the light of the above arguments, how to choose the appropriate representation of the human decisions and behaviour to be used in the agent-based modelling of demographic processes? First, we need to remember that agent-based models are a very convenient tool, with which we can integrate behavioural theory with social theory and data. This implies that – as suggested in the Manifesto of computational social science \citep{Conte} – such models would ideally need to collect additional information on human decision making through bespoke cognitive experiments, instead of merely relying on hypotheses or assumptions \citep{Courgeau}. In this way, the choice of a particular choice model will become an empirical issue, but specific to a given context and the decision problem of interest. Following this approach would allow for including empirical micro-foundations in computational demographic models in an open and transparent way – and provide a natural, bottom-up and empirical way if evaluating the suitability of particular choice theories for the problem at hand.
